`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-02-10 paper/paper.qmd
extra examples of potential applications of the package

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations. We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], or understanding access to agricultural processing facilities [@zhao2023replanting].

@@

`traveltime` provides a spatial interface using object classes from the `terra` package [@terra]. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object. Internally, `traveltime` uses a number of existing R packages to perform specific operations. `gdistance` [@gdistance2017] is used to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over any 'friction surface' the user supplies. To provide easy access to the existing friction surfaces generated by @weiss2020global. `traveltime` uses the package `malariaAtlas` to access friction surfaces for the desired extent, potentially saving the user downloading global surfaces and cropping them separately.

>>>>

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations. We envisage this software having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], allocating public defibrillators [tierney2018novel], setting rehabilitation districts for stroke patients [padgham2019introduction], or understanding access to agricultural processing facilities [@zhao2023replanting].

@@

`traveltime` provides a spatial interface using object classes from the `terra` package [@terra]. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object. The travel time is calculated as movement over a resistance 'friction surface' [@gdistance2017]. To provide easy access to the existing friction surfaces generated by @weiss2020global. `traveltime` uses the package `malariaAtlas` to access friction surfaces for the desired extent. Users can also supply other friction surfaces. Functionality from `gdistance` [@gdistance2017] is also used internally to calculate the minimum least-cost-distance for each cell from the points of interest.
